---
title: 'R Supplament 1: Working with R'
author: "Mhairi McNeill and Robert Raeside"
date: "17 June 2016"
output: word_document
---

To analyse numeric and categorical data we are going to use a programming language called R. You might have used a point-and-click interface for analysing data before like Excel, SPSS or Minitab. Using R is somewhat similar to these programmes but also somewhat different. The biggest difference is that you'll be typing commands rather than clicking on boxes and menus. At first this is more difficult but will almost certainly become quicker and easier in the long run. 

There's a bit of debate about what software you should use to analyse data. Every way of doing it has advantages and disadvantages. We'll be using R in this course because:

  1. When you type everything you have a written record of everything you've done to a piece of data. This makes it easier to check your logic and the preserve the original data. 
  2. You will often need to do the same piece of analysis over and over on different datasets. Having the code to do the analysis written out makes it easy to repeat.
  3. R is free and open source. 
  4. R has a lot of excellent libraries (downloadable extensions to R) which make generating plots, doing specialised techniques, making reports and even producing interactive dashboards easy. 

If you've learnt to programme before, that's great. I'm sure you will find learning R straightforward. There are a few idiosyncratic features of R that can trip up experienced programmers. You might find this guide useful: http://www.johndcook.com/blog/r_language_for_programmers/

If you've never done programming before that is also great. R is a good language for a beginner and once you've learnt R it will be much easier to pick up other languages, which opens up a huge range of things you can do with computers! 

### Installing R

We are going to install R the programming language. R is programme that actually does the analysis - you interact with this programme by typing R commands. For example, if you type `2 + 2`, R will do the maths and return the result to you. 

Then we are going to install an IDE (interactive development environment) called RStudio. RStudio is a very popular way of interacting with R. While you don't need it to write R code RStudio makes it easy to do things like:
1. Write longer pieces of code and get R to run it piece by piece.
2. Keep track of what data R knows about.
3. Organise all your R code files.
4. See plots and tables you've created in R.
5. Make special types of R files e.g. reports, dashboards etc.
6. Access R's in-built help
7. Manage which R packages (extensions to R) you have installed. 

R is free and open source. It is written by volunteers and all the all packages you'll use were also written by volunteers.

RStudio is also free and open source, but is made by a profit making company. They make their money by selling a professional version of RStudio that runs on a sever and has support.

[Videos here]

### Getting started.

Once you have RStudio installed and opened you should see four panels. We'll ignore the two right-hand-side panels for just now. 

The bottom-left panel is the R interpreter. We can type R commands into the interpreter and R will do some calculations and return an answer. Let's start with one of the most basic commands possible. Let's get R to add 2 and 2. Type `2 + 2` into the interpreter and press enter. 

You should see something like this:

```{r}
2 + 2
```

After you hit enter R understood the command and found the answer, returning it almost instantly to you. (The little [1] in front of the answer just means that there's only one element to the answer).

Now, type the same thing in the top left panel. When you press enter here nothing will happen. The top left panel is basically just a very simple text editor like Notepad on windows or TextEdit on a Mac. You can absolutely write R code in a separate text editor and many people do. A big advantage of writing code inside RStudio is that it's very easy to transfer code from the editor to the R interpreter. Just move your cursor to the line with `2 + 2` and press ctrl and enter at the same time. Pressing cmd+enter also works on a Mac. The code you have written will now appear in the interpreter along with the answer.

If you have a longer piece of code that goes across multiple lines you will need to highlight the lines and then press ctrl+enter. Try this just now after typing this in the top left editor.

```{r}
1 + 2 +
4
```

You can save the code you have written in the text editor and come back to it at any time. Just go to **File** and then **Save As…**. When going through this unit keep everything you’ve written in the text editor saved. 
 
### Vectors

A very important thing to understand about R is that it works with *vectors*. Vectors are a list of items. Often this will be a list of numbers, but we will also work with vectors of strings (in programming languages anything made up of letters is called a string) and vectors made up of True and False values. In R we call a vector of strings a character vector. A vector made up of only True and False is a logical vector. 

There are other types of vector in R, but for just now we will only be working with numeric, character and logical vectors. To make a vector in R we surround the elements with `c(` and `)`. For example

```{r}
c(1, 4, 6, 3)
```

Below is an example of the three vectors types we will learn about.

Numeric   | `c(2.2, 0.1, 5)`
Character | `c('apple', 'pear', 'e')`
Logical   | `c(TRUE, FALSE, FALSE)`

### Exercise

1. Create a numeric vector containing this data: 10, 11, 11, 14. (Type in the text editor and use ctrl+enter to send to the interpreter)
2. Create a character vector containing this data: 'small', 'small', 'small', 'large'.

### Assignment

Creating a vector isn't very useful if we can't store that vector somewhere. In a programming language if we want to store some data, or the results of a calculation we use assignment. Assignment is basically giving a name to a piece of data. If we want to use that data we can just find it by using its name, rather than typing out the data again. 

In R we can assign using an arrow. Here’s an example of assigning the name x to a vector:

```{r}
x <- c(1, 4, 6, 3)
```

Now we've assigned a vector to x we can see what x is at any point by just typing x into the interpreter.

```{r}
x
```

Here we have been giving names to vectors but we can give a name to any type of R object. Here's how we would assign a name to a number:

```{r}
days <- 5
```

We can also give names to more complicated objects which will will see later.

Names in R can only contain letters, numbers, '.' and '_'. They can't start with a number or an underline. Try to make names descriptive, so that when you go back to a piece of analysis you should be able to understand what everything is. Since you can't have spaces in names you can use the underscore character as a space: 

```{r}
days_until_launch <- 5
```

### Exercise

1. Give the numeric vector we created earlier the name `height`.
2. Give the character vector the name `height_category`.

### The difference between 'apple' and apple

A very common point of confusion for new programmers is the difference between a string like `'apple'` and a name like `apple`. 

First first `'apple'` with quotes is simply a piece of data. If we type this into the interpreter R sees a new string we have created.

```{r}
'apple'
```

The second is a name given to an object. Since we haven’t called anything `apple` yet R doesn't know anything about `apple`.

```{r, error = TRUE}
apple
```

Another way of putting it is that `'apple'` is something that gets given a name:

```{r}
fruit <- 'apple'
```

And `apple` is a name that you can give to data:

```{r}
apple <- TRUE
```

You might also see people make names like this:
  
  ```{r}
daysUntilLaunch <- 5
```

Which is totally fine, as long as you are consistent in your naming style. Some R users create names like this:
  
  ```{r}
days.until.launch <- 5
```

Which is again fine, although less recommended because many other languages won't let you use `.` in a name, so that could be potentially confusing for people reading your code who know other languages.  


Now you've understood the basics of R, we can begin analysing some data. Starting with some very basic, small data we'll discover ways the data can visualised and summarised which will help us understand it better and answer questions we have about the data. 


